---
output: 
  html_document: 
    highlight: default
    theme: spacelab
---



<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{VP Misc}
%\VignetteKeyword{vp.mics}
%\VignettePackage{vp.misc}
-->

```{r setup, message=FALSE, warning=FALSE}
# set global chunk options
library(knitr)
library(dplyr)
opts_chunk$set(size='tiny')
```

# Purpose
To show different way to convert from one type of gene/protein IDs to another.
The example task is to get UniProt [accession](http://www.uniprot.org/help/accession_numbers) 
and [entry names](http://www.uniprot.org/help/entry_name) based on
gene symbols. Bioconductor has multiple ways to get this done.
An example gene list.
```{r init}
genes <- c("TH","GFAP","CLU","SNCA","APP","MAPT")
```

# UniProt.ws package
There is now a key that directly corresponds to gene symbols. 
However, GENECARDS ID works well for this purpose.
```{r uniprot, warning=FALSE, message=FALSE}
library(UniProt.ws)
up <- UniProt.ws(taxId=9606)
select(up, keys=genes, columns=c("ENTRY-NAME","UNIPROTKB"), keytype="GENECARDS")
```

# org.Hs.eg.db package
The OrgDb packages are centered around Entrez IDs. Thus, first gene symbols
need to be converted to Entrez IDs and then to UniProt IDs. The only UniProt IDs
available through OrbDb are accessions. An important note is that it contains
both reviewed and unreviewed accessions. This can cause a trouble once in a while.
For example it will be confusing which accession to pick for the `APP` gene 
(at least in the `2017_05` UniProt release).
```{r orgDb, warning=FALSE, message=FALSE}
library(org.Hs.eg.db)
entrez_ids <- org.Hs.egSYMBOL2EG[genes] # AnnDbBimap object
# selecting first Entrez ID in case there are multiple
entrez_ids <- sapply(as.list(entrez_ids), '[', 1)
# selecting first accession in case there are multiple. 
# This is sometimes necessary, but dangerous step as the first one may not be
# the primary ID.
uniprot_ids <- sapply(as.list(org.Hs.egUNIPROT[entrez_ids]), '[', 1)
data.frame(genes, uniprot_ids)
```
alternative
```{r orgDb2, message=FALSE}
entrez_ids <- org.Hs.egSYMBOL2EG[genes] %>% as.data.frame
uniprot_ids <- org.Hs.egUNIPROT[entrez_ids$gene_id] %>% as.data.frame
inner_join(entrez_ids, uniprot_ids)
```


# clusterProfiler package
clusterProfiler has a convenience function bitr convienience function that at
the backend uses OrgDb. Thus the problem of carrying through unreviewed IDs is 
inherited from OrgDb.
```{r clusterProfiler, message=FALSE}
library(clusterProfiler)
bitr(genes, fromType="SYMBOL", toType="UNIPROT", OrgDb="org.Hs.eg.db")
```


# biomaRt approach
This one may or may not to work.  There some persistent troubles due to [BioMart 
site migration](http://www.biomart.org/news.html).  
When it works, it is really great, though.
```{r biomartm, message=FALSE}
library(biomaRt)
# listMarts()
ensembl <- useMart("ensembl")
# listDatasets(ensembl)
hsa <- useDataset( "hsapiens_gene_ensembl", mart=ensembl)
# listAttributes(hsa)
# listFilters(hsa)
getBM(attributes= c("hgnc_symbol","uniprotswissprot"), 
              filters=c("hgnc_symbol"), 
              values=genes, 
              mart=hsa)
```


