---
output: 
  html_document: 
    highlight: default
    theme: spacelab
---



<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{ontology overrepresentation}
%\VignetteKeyword{vp.mics}
%\VignettePackage{vp.misc}
-->

```{r setup}
# set global chunk options
library(knitr)
opts_chunk$set(size='tiny')
```

# Purpose
The whole purpose of this vignette/cheatsheet is to show different approaches
for testing for significance of certain over-representation in one group of 
genes vs another.

# The data 
we'll take a look what is enriched in the sixth cluster compare to the rest
```{r libs, results='hide', message=FALSE, warning=FALSE}
library(vp.misc)
library(clusterProfiler)
library(GOstats)
library(ReactomePA)
```
```{r init2}
data(gcSample)
# there are 8 clusters
sapply(gcSample, length)
# the IDs are Entrez Gene IDs
head(gcSample[[1]])
```


***


# KEGG annotation
## `GOstats` package
```{r kegg_gostats}
p <- new("KEGGHyperGParams",
     geneIds=gcSample[[6]],
     universeGeneIds=unique(unlist(gcSample)),
     pvalueCutoff = 1,
     annotation="org.Hs.eg.db",
     testDirection="over")
xke1 <- hyperGTest(p)
res <- summary(xke1)[,c("KEGGID","Term","Pvalue","Count","Size")]
kable(head(res))
```

## `clusterProfiler` package
```{r kegg_clusterprofiler}
xke2 <- enrichKEGG(gcSample[[6]], universe = unlist((gcSample)),
                   pvalueCutoff = 1, qvalueCutoff = 1,
                   minGSSize = 0,
                   use_internal_data=TRUE) # then uses KEGG.db as GOstats below
res <- summary(xke2)[,c("ID","Description","pvalue","GeneRatio","BgRatio")]
rownames(res) <- NULL
kable(head(res))
```


***


# GO BP annotation
Pretty much same thing as above, but with a few tweaks 
## `GOstats` package
```{r gobp_gostats}
p <- new("GOHyperGParams",
     geneIds=gcSample[[6]],
     universeGeneIds=unique(unlist(gcSample)),
     pvalueCutoff = 1,
     annotation="org.Hs.eg.db",
     ontology="BP",
     conditional=FALSE, # influential option
     testDirection="over")
xgobp1 <- hyperGTest(p)
res <- summary(xgobp1)[,c("GOBPID","Term","Pvalue","Count","Size")]
kable(head(res))
```

## `clusterProfiler` package
```{r gobp_clusterprofiler}
xgobp2 <- enrichGO(gcSample[[6]], 
                   OrgDb = "org.Hs.eg.db",
                   universe = unique(unlist((gcSample))),
                   ont = "BP", pvalueCutoff = 1, qvalueCutoff = 1,
                   minGSSize = 0) 
xgobp2 <- subset_by_size(xgobp2, maxObsSize = 50)
res <- summary(xgobp2)[,c("ID","Description","pvalue","GeneRatio","BgRatio")]
rownames(res) <- NULL
kable(head(res))
```

There is a discrepancy between the results.

clusterProfiler does the test in the following chain:

`enrichGO -> DOSE:::enrich.internal`

GOstats chain:

`hyperGTest -> GOstats:::.hyperGTestInternal -> Category:::.doHyperGTest -> Category:::.doHyperGInternal`

It seems that GOstats does quite a bit of gene and 
GO term removal prior the test.


***


# GO CC annotation
## `GOstats` package
```{r gocc_gostats}
p <- new("GOHyperGParams",
     geneIds=gcSample[[6]],
     universeGeneIds=unique(unlist(gcSample)),
     pvalueCutoff = 1,
     annotation="org.Hs.eg.db",
     ontology="CC",
     conditional=FALSE, # influential option
     testDirection="over")
xgocc1 <- hyperGTest(p)
res <- summary(xgocc1)[,c("GOCCID","Term","Pvalue","Count","Size")]
kable(head(res))
```

## `clusterProfiler` package
```{r gocc_clusterprofiler}
xgocc2 <- enrichGO(gcSample[[6]], 
                   OrgDb = "org.Hs.eg.db",
                   universe = unique(unlist((gcSample))),
                   ont = "CC", pvalueCutoff = 1, qvalueCutoff = 1,
                   minGSSize = 0) 
xgocc2 <- subset_by_size(xgocc2, maxObsSize = 50)
res <- summary(xgocc2)[,c("ID","Description","pvalue","GeneRatio","BgRatio")]
rownames(res) <- NULL
kable(head(res))
```


***


# GO MF annotation
## `GOstats` package
```{r gomf_gostats}
p <- new("GOHyperGParams",
     geneIds=gcSample[[6]],
     universeGeneIds=unique(unlist(gcSample)),
     pvalueCutoff = 1,
     annotation="org.Hs.eg.db",
     ontology="MF",
     conditional=FALSE, # influential option
     testDirection="over")
xgomf1 <- hyperGTest(p)
res <- summary(xgomf1)[,c("GOMFID","Term","Pvalue","Count","Size")]
kable(head(res))
```

## `clusterProfiler` package
```{r gomf_clusterprofiler}
xgomf2 <- enrichGO(gcSample[[6]], 
                   OrgDb = "org.Hs.eg.db",
                   universe = unique(unlist((gcSample))),
                   ont = "MF", pvalueCutoff = 1, qvalueCutoff = 1,
                   minGSSize = 0) 
xgomf2 <- subset_by_size(xgomf2, maxObsSize = 50)
res <- summary(xgomf2)[,c("ID","Description","pvalue","GeneRatio","BgRatio")]
rownames(res) <- NULL
kable(head(res))
```


***

# Reactome
As for _Reactome_ there is only one pre-canned solution - from `ReactomePA`

## `ReactomePA` package
```{r reactome_clusterprofiler}
xreac <- enrichPathway(gcSample[[6]], 
                       organism = "human",
                       universe = unique(unlist((gcSample))),
                       pvalueCutoff = 1, qvalueCutoff = 1, minGSSize = 0) 
xreac <- subset_by_size(xreac, maxObsSize = 50)
res <- summary(xreac)[,c("ID","Description","pvalue","GeneRatio","BgRatio")]
rownames(res) <- NULL
kable(head(res))
```

There is a warning because some genes map to both human and mouse pathways.
Later this creates trouble. Here is my [post](https://support.bioconductor.org/p/70921/) about that on
Bioconductor's mailing list.

<!--
The problem is that DOSE:::enrich.internal in its line
qExtID2TermID = EXTID2TERMID(gene, organism, ...)
returns pathway mapping mouse. Then obviously it can not
fetch description of mouse pathway from human stuff here
Description <- TERM2NAME(qTermID, organism, ...)
debugonce(DOSE::enrich.internal)
By the way, "organism" argument does not do anything in
qExtID2TermID = EXTID2TERMID(gene, organism, ...)
-->

***

# PFAM
## `GOstats` package
```{r pfam_gostats}
p <- new("PFAMHyperGParams",
     geneIds=gcSample[[6]],
     universeGeneIds=unique(unlist(gcSample)),
     pvalueCutoff = 1,
     annotation="org.Hs.eg.db",
     testDirection="over")
xpfam <- hyperGTest(p)
res <- summary(xpfam)
kable(head(res))
```

Adding description to PFAM term is something to think about.


***

# Custom stuff

Taking advantage of `enricher` function from `clusterProfiler`

```{r custom_enricher}
# randomly assign genes to 20 clusters
setnum <- replicate(length(unique(unlist(gcSample))), sample(1:20, 1))
setnum <- paste("set", setnum, sep = '')
TERM2GENE <- data.frame(term=setnum, gene=unique(unlist(gcSample)),
                        stringsAsFactors = FALSE)
TERM2NAME <- data.frame(term=setnum, 
                        name=paste("Description: ", setnum, sep = ''),
                        stringsAsFactors = FALSE)
xfake <- enricher(gene = gcSample[[6]], 
                  universe = unique(unlist(gcSample)),
                  pvalueCutoff = 1, qvalueCutoff = 1, minGSSize = 0,
                  TERM2GENE = TERM2GENE, 
                  TERM2NAME = TERM2NAME)
res <- summary(xfake)[,c("ID","Description","pvalue","GeneRatio","BgRatio")]
rownames(res) <- NULL
kable(head(res))
```
```{r show_fake}
cnetplot(xfake, vertex.label.cex = 0.5)
```

A different layout
```{r show_fake2}
cnetplot(xfake, vertex.label.cex = 0.5, layout="kk")
```

Check out as well [HTSanalyzeR](http://bioconductor.org/packages/release/bioc/html/HTSanalyzeR.html) and  [geecc](http://bioconductor.org/packages/release/bioc/html/geecc.html) Bioconductor packages.






