---
title: "Over-representation Analysis"
output: 
  html_document: 
    highlight: default
    theme: spacelab
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{ontology overrepresentation}
%\VignetteKeyword{MSnSet.utils}
%\VignettePackage{MSnSet.utils}
-->

```{r setup}
# set global chunk options
library(knitr)
opts_chunk$set(size='tiny')
```

# Purpose
The whole purpose of this vignette/cheatsheet is to show different approaches
for testing for significance of certain over-representation in one group of 
genes vs another.

# The data 

we'll take a look what is enriched in the sixth cluster compare to the rest

```{r libs, results='hide', message=FALSE, warning=FALSE}
library(clusterProfiler)
library(GOstats)
library(ReactomePA)
library(org.Hs.eg.db)
library(enrichplot)
```

```{r init2}
data(gcSample)
# there are 8 clusters
sapply(gcSample, length)
# the IDs are Entrez Gene IDs
head(gcSample[[1]])
```


***


# KEGG annotation
## `GOstats` package
```{r kegg_gostats}
p <- new("KEGGHyperGParams",
         geneIds = gcSample[[6]],
         universeGeneIds = unique(unlist(gcSample)),
         annotation = "org.Hs.eg.db",
         pvalueCutoff = 1,
         testDirection = "over")
xke1 <- hyperGTest(p)
res <- summary(xke1)[, c("KEGGID", "Term", "Pvalue", "Count", "Size")]
kable(head(res))
```

## `clusterProfiler` package
```{r kegg_clusterprofiler}
xke2 <- enrichKEGG(gene = gcSample[[6]], 
                   organism = "hsa",
                   keyType = "kegg",
                   pvalueCutoff = 1,
                   pAdjustMethod = "BH",
                   universe = unique(unlist((gcSample))),
                   minGSSize = 0,
                   qvalueCutoff = 1,
                   use_internal_data = FALSE)
res <- xke2@result[, c("ID", "Description", "pvalue", "GeneRatio", "BgRatio")]
rownames(res) <- NULL
kable(head(res))
```


***


# GO Biological Process (BP) annotation

Pretty much same thing as above, but with a few tweaks.

## `GOstats` package

```{r gobp_gostats}
p <- new("GOHyperGParams",
         ontology = "BP", # Gene Ontology: Biological Process
         conditional = FALSE, # Should calculation condition on GO structure?
         geneIds = gcSample[[6]],
         universeGeneIds = unique(unlist(gcSample)),
         annotation = "org.Hs.eg.db",
         pvalueCutoff = 1,
         testDirection = "over")
xgobp1 <- hyperGTest(p)
res <- summary(xgobp1)[, c("GOBPID", "Term", "Pvalue", "Count", "Size")]
kable(head(res))
```

## `clusterProfiler` package

```{r gobp_clusterprofiler}
xgobp2 <- enrichGO(gene = gcSample[[6]], 
                   OrgDb = "org.Hs.eg.db",
                   ont = "BP",
                   pvalueCutoff = 1,
                   universe = unique(unlist((gcSample))),
                   qvalueCutoff = 1,
                   minGSSize = 0)
# xgobp2 <- subset_by_size(xgobp2, maxObsSize = 50)
res <- xgobp2@result[, c("ID", "Description", "pvalue", "GeneRatio", "BgRatio")]
rownames(res) <- NULL
kable(head(res))
```

***

# GO Cellular Component (CC) annotation

## `GOstats` package

```{r gocc_gostats}
p <- new("GOHyperGParams",
         ontology = "CC", # Gene Ontology: Cellular Component
         conditional = FALSE, # Should calculation condition on GO structure?
         geneIds = gcSample[[6]],
         universeGeneIds = unique(unlist(gcSample)),
         annotation = "org.Hs.eg.db",
         pvalueCutoff = 1,
         testDirection = "over")
xgocc1 <- hyperGTest(p)
res <- summary(xgocc1)[, c("GOCCID", "Term", "Pvalue", "Count", "Size")]
kable(head(res))
```

## `clusterProfiler` package

```{r gocc_clusterprofiler}
xgocc2 <- enrichGO(gene = gcSample[[6]], 
                   OrgDb = "org.Hs.eg.db",
                   ont = "CC",
                   pvalueCutoff = 1,
                   universe = unique(unlist((gcSample))),
                   qvalueCutoff = 1,
                   minGSSize = 0)
# xgocc2 <- subset_by_size(xgocc2, maxObsSize = 50)
res <- xgocc2@result[, c("ID", "Description", "pvalue", "GeneRatio", "BgRatio")]
rownames(res) <- NULL
kable(head(res))
```

***

# GO Molecular Function (MF) annotation

## `GOstats` package

```{r gomf_gostats}
p <- new("GOHyperGParams",
         ontology = "MF", # Gene Ontology: Molecular Function
         conditional = FALSE, # Should calculation condition on GO structure?
         geneIds = gcSample[[6]],
         universeGeneIds = unique(unlist(gcSample)),
         annotation = "org.Hs.eg.db",
         pvalueCutoff = 1,
         testDirection = "over")
xgomf1 <- hyperGTest(p)
res <- summary(xgomf1)[, c("GOMFID", "Term", "Pvalue", "Count", "Size")]
kable(head(res))
```

## `clusterProfiler` package

```{r gomf_clusterprofiler}
xgomf2 <- enrichGO(gcSample[[6]], 
                   OrgDb = "org.Hs.eg.db",
                   universe = unique(unlist((gcSample))),
                   ont = "MF", pvalueCutoff = 1, qvalueCutoff = 1,
                   minGSSize = 0) 
# xgomf2 <- subset_by_size(xgomf2, maxObsSize = 50)
res <- xgomf2@result[, c("ID", "Description", "pvalue", "GeneRatio", "BgRatio")]
rownames(res) <- NULL
kable(head(res))
```


***

# Reactome
As for _Reactome_ there is only one pre-canned solution - from `ReactomePA`

## `ReactomePA` package

```{r reactome_clusterprofiler}
xreac <- enrichPathway(gene = gcSample[[6]], 
                       organism = "human",
                       pvalueCutoff = 1,
                       qvalueCutoff = 1,
                       universe = unique(unlist((gcSample))),
                       minGSSize = 0)
# xreac <- subset_by_size(xreac, maxObsSize = 50)
res <- xreac@result[, c("ID", "Description", "pvalue", "GeneRatio", "BgRatio")]
rownames(res) <- NULL
kable(head(res))
```

***

# PFAM

## `GOstats` package

```{r pfam_gostats}
p <- new("PFAMHyperGParams",
         geneIds = gcSample[[6]],
         universeGeneIds = unique(unlist(gcSample)),
         annotation = "org.Hs.eg.db",
         pvalueCutoff = 1,
         testDirection = "over")
xpfam <- hyperGTest(p)
res <- head(summary(xpfam)) # Top 6

kable(res)
```


***

# Custom stuff

Taking advantage of `enricher` function from `clusterProfiler`

```{r custom_enricher}
# randomly assign genes to 20 clusters
set.seed(1236)
setnum <- replicate(length(unique(unlist(gcSample))), sample(1:20, 1))
setnum <- paste("set", setnum, sep = '')
TERM2GENE <- data.frame(term = setnum, gene = unique(unlist(gcSample)),
                        stringsAsFactors = FALSE)
TERM2NAME <- data.frame(term = setnum, 
                        name = paste("Description: ", setnum, sep = ''),
                        stringsAsFactors = FALSE)
xfake <- enricher(gene = gcSample[[6]], 
                  pvalueCutoff = 1,
                  universe = unique(unlist(gcSample)),
                  minGSSize = 0,
                  qvalueCutoff = 1,
                  TERM2GENE = TERM2GENE, 
                  TERM2NAME = TERM2NAME)
res <- xfake@result[, c("ID", "Description", "pvalue", "GeneRatio", "BgRatio")]
rownames(res) <- NULL
kable(head(res))
```

```{r show_fake}
cnetplot(xfake, layout = "fr", cex_label_category = 0.8, cex_label_gene = 0.4)
```

A different layout

```{r show_fake2}
cnetplot(xfake, layout = "kk", cex_label_category = 0.8, cex_label_gene = 0.4)
```

