---
title: "PeCorA" 
output: 
  html_document: 
    highlight: tango
---

<!--
%\VignetteEngine{knitr::rmarkdown} 
%\VignetteIndexEntry{PeCorA}
%\VignetteKeyword{MSnSet.utils} 
%\VignettePackage{MSnSet.utils}
-->

```{r setup}
# set global chunk options
library("knitr")
opts_chunk$set(size='tiny')
```

```{r env, message=FALSE, echo=FALSE, cache=FALSE, warning=FALSE}
library(MSnSet.utils)
library(ggplot2)

# Example MSnSet
data("brain_msnset")
m <- brain_msnset # shorter object name for vignette
m # display summary
head(fData(m))
```

PeCorA will estimate the significance of interaction between the peptides
mapping to a protein and the treatment variable supplied.

```{r}
# proteins with peptides to test
proteins <- c("APOE", "BIN1")

# Categorical treatment
m <- PeCorA(m,
            treatment = "var_y",
            control_group = "A",
            proteins = proteins)

# Show fData for those proteins
head(fData(m)[fData(m)[["Protein"]] %in% proteins, ])
```

Running PeCorA again will add new results to fData.

```{r}
# Numeric treatment
m <- PeCorA(m,
            treatment = "var_x",
            proteins = proteins,
            median_mod = FALSE)

# Display part of results
head(fData(m)[fData(m)[["Protein"]] %in% proteins, ])
```

```{r}
# Update results - use median_mod = TRUE this time
m <- PeCorA(m,
            treatment = "var_x",
            proteins = proteins,
            median_mod = TRUE)

# Display part of results
head(fData(m)[fData(m)[["Protein"]] %in% proteins, ])
```

Finally, `plot_PeCorA` can plot a particular feature (peptide) against all other
peptides mapping to the same protein, and will include the adjusted p-value for
that peptide in the plot subtitle. Depending on the type of the treatment
variable (categorical vs. numeric) a boxplot or scatterplot is made,
respectively. The parameter `median_mod` can be set to `TRUE` to aggregate
(samplewise) all other peptides mapping to the same protein using the median.

```{r}
# Feature (peptide) to plot
peptide <- "VFEEMNVDLQEELPSLWNSR"
protein <- "BIN1"

# Numeric treatment - scatterplot
p1 <- plot_PeCorA(m, 
                  peptide = peptide, 
                  protein = protein,
                  treatment = "var_x", 
                  median_mod = FALSE)
p1
```

```{r}
# Categorical treatment - boxplots
p2 <- plot_PeCorA(m, 
                  peptide = peptide, 
                  protein = protein,
                  treatment = "var_y", 
                  median_mod = FALSE)
p2
```

Since this is a `ggplot` object, we can modify it as needed.

```{r}
# Change y-axis limits and theme
p3 <- p2 + 
  scale_y_continuous(limits = c(-6, 8),
                     expand = expansion(mult = 5e-3)) +
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.position = "bottom")
p3
```


