---
title: "Example of WGCNA Analysis"
subtitle: "Using Ovarian Cancer CPTAC Dataset as Example"
output: 
  BiocStyle::html_document: 
    toc: true
    number_sections: true
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{WGCNA example}
%\VignetteKeyword{vp.mics}
%\VignettePackage{vp.misc}
-->


```{r setup, echo=FALSE}
# set global chunk options
library(knitr)
opts_chunk$set(size='tiny')
```

# Purpose
Example of WGCNA analysis. The package is available on [CRAN](https://cran.r-project.org/web/packages/WGCNA/index.html). The tutorial 
available at [UCLA web-site](https://labs.genetics.ucla.edu/horvath/CoexpressionNetwork/Rpackages/WGCNA/). Here we use our own ovarian cancer CPTAC data set.

# Loading
## Libraries
```{r init, message=FALSE, warning=FALSE, results='hide'}
library(WGCNA)
options(stringsAsFactors = FALSE)
enableWGCNAThreads()
disableWGCNAThreads() # on RStudio for some reason multithreading does not work
library(vp.misc)
```

## Data as MSnSet object

```{r data}
data("cptac_oca")
```

WGCNA seems to be tolerant to missing values, but we subset the 
data for the sake of faster computation.
```{r data2}
oca.set <- oca.set[complete.cases(exprs(oca.set)),]
```



# Selecting Transform Power
Scanning powers to find the ones that produce scale-free topology. 
```{r power_scan}
powers = 1:20
sft <- pickSoftThreshold(t(exprs(oca.set)),
                         powerVector = powers,
                         networkType = "signed",
                         verbose = 0)
```

```{r plotting_power_scan, fig.width=8, fig.height=4, fig.align='center', out.width='800px'}
par(mfrow = c(1,2))
cex1 = 0.9
plot(sft$fitIndices[,1],
     -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab="Soft Threshold (power)",
     ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"))
text(sft$fitIndices[,1],
     -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=cex1,col="red")
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")
# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1],
     sft$fitIndices[,5],
     xlab="Soft Threshold (power)",
     ylab="Mean Connectivity",
     type="n",
     main = paste("Mean connectivity"))
text(sft$fitIndices[,1],
     sft$fitIndices[,5],
     labels=powers,
     cex=cex1,
     col="red")
```

The lowest power that results in crossing in 0.9 threshold is 10. Therefore we'll
set it as a final value.
```{r soft_threshold}
softPower = 10
```


# Clustering. First Round.
The recommended type of correlation is signed. That way correlations +1
and -1 will have different meaning. In case of unsigned correlation the 
sign obviosly does not matter and genes with +1 correlation as close to 
each other as -1 correlation. In case of (recommended) signed correlation
the adjacency (or 1 - distance) is calculated as
`adjacency = (0.5*(1+cor))^power`.
```{r adjacency, cache=TRUE}
adjacency.mat <- adjacency(t(exprs(oca.set)), power = softPower, type="signed")
```
The gene-gene adjacencies converted to topological overlap. This helps to 
"smooth" correlation structure as it relies not only direct gene-gene 
correlations, but also the correlations with genes that are connected in common.
```{r topological_overlap}
TOM <- TOMsimilarity(adjacency.mat, TOMType = "unsigned")
```
Finally adjacency or its analog, topological overlap, converted to distance.
```{r to_distance}
dissTOM = 1-TOM
```
Calling base R `hclust` function to perform hierarchical clustering.
```{r clustering, cache=TRUE}
geneTree <- hclust(as.dist(dissTOM), method = "average")
```
Plot the resulting clustering tree
```{r plot_tree_1}
plot(geneTree, xlab="", sub="", 
     main = "Gene clustering on TOM-based dissimilarity",
     labels = FALSE, hang = 0.04)
```


# Dynamic Tree Cut
Cutting the tree to identify the modules.
```{r dynamic_tree_cut}
minModuleSize <- 20 # 20 is default
dynamicMods <- cutreeDynamic(dendro = geneTree,
                             distM = dissTOM,
                             # deepSplit = 2, 
                             # pamRespectsDendro = FALSE,
                             # method= "tree",
                             minClusterSize = minModuleSize)
```


# Coloring The Modules
As it is evident from the commented code, it is possible to define coloring in
multiple ways. It is practical to take advantage of WGCNA's `labels2colors`
because they provide human-readable names to the modules. Although 
leveraging `RColorBrewer` may provide better colors, however the names
will be hex color codes, which makes invonvenient naming.
```{r coloring_modules}
# Convert numeric lables into colors
library(RColorBrewer)
# dynamicColors <- labels2colors(dynamicMods, colorSeq = brewer.pal(11,'Spectral'))
# this `standardColors()` is a bit nicer as the colors come with names
dynamicColors <- labels2colors(dynamicMods, colorSeq = standardColors())
# Plot the dendrogram and colors underneath
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    main = "Gene dendrogram and module colors")
```



# Decision on Module Merger
Modules can be merged if they are too similar. The similarity is quantified as 
correlation between the eigengenes. Threshold for module merged visualized
on the dendrogram.

Calculating eigengenes.
```{r calculating_eigengenes}
MEList <- moduleEigengenes(t(exprs(oca.set)), colors = dynamicColors)
MEs <- MEList$eigengenes
```

Calculate distances between module eigengenes. This is equivalent to 
signed correlation distance. That is +1 correlation is zero distance or
very similar modules, while -1 are the furthest modules possible.
```{r module_distances}
MEDiss <- 1 - cor(MEs)
```

Cluster module eigengenes to find out which are to be merged (assuming
the distance is no larger then a predefined threshold).
```{r eigengene_clustering}
METree <- hclust(as.dist(MEDiss), method = "average")
```

Plot the cut line into the dendrogram.
```{r plot_eigengene_clustering}
plot(METree, main = "Clustering of module eigengenes",
     xlab = "", sub = "")
MEDissThres <- 0.5 # default value is 0.2. 0.5 here is for demo purpose
abline(h=MEDissThres, col = "red")
```


# Module Merger
Call an automatic merging function. The threshold is defined above 
`MEDissThres = ` ``r MEDissThres``. These are going to be the final modules.
```{r module_merge}
merge <- mergeCloseModules(t(exprs(oca.set)),
                           dynamicColors, 
                           cutHeight = MEDissThres, 
                           verbose = 3)
moduleColors <- merge$colors
```

Show both original and merged modules.
```{r showing_modules_after_merger}
plotDendroAndColors(geneTree, cbind(dynamicColors, moduleColors),
                    c("Dynamic Tree Cut", "Merged dynamic"),
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    marAll = c(1, 7, 3, 1))
```

Recalculating eigengenes.
```{r recalc_eigengenes}
MEs0 <- moduleEigengenes(t(exprs(oca.set)), moduleColors)$eigengenes
MEs <- orderMEs(MEs0)
```



# Module-Trait Association

Preparing traits data.
```{r prepare_traits}
datTraits <- pData(oca.set)[,grep('binary', varLabels(oca.set))]
colnames(datTraits) <- sub('.binary','',colnames(datTraits))
head(datTraits)
```

Calculating correlations and associated p-values.
```{r calculating_correlations_pvalues}
moduleTraitCor = cor(MEs, datTraits, use = "p");
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, ncol(oca.set));
```

Displaying correlations and their p-values as a heatmap.
```{r trait_association_heatmap}
textMatrix <-  paste(signif(moduleTraitCor, 2),
                     "\n(",
                     signif(moduleTraitPvalue, 1),
                     ")", sep = "");
textMatrix <- matrix(textMatrix, ncol = ncol(moduleTraitCor))
par(mar = c(6, 8, 3, 2))
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = colnames(datTraits),
               yLabels = names(MEs),
               ySymbols = names(MEs),
               colorLabels = FALSE,
               # colors = greenWhiteRed(50),
               colors = blueWhiteRed(50),
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 0.75,
               zlim = c(-1,1),
               main = paste("Module-Trait Association (p-value)"))
```


# Ontology Enrichment. Reactome.
```{r ontology_enrichment, fig.height=8, fig.width=16, message=FALSE, warning=FALSE}
# split by module color
features <- sub('\\.\\d+','',featureNames(oca.set))
clustList <- tapply(features, moduleColors, c)
library(org.Hs.eg.db)
clustList <- lapply(clustList, intersect, mappedRkeys(org.Hs.egREFSEQ2EG))
clustEG <- lapply(clustList, function(x) sapply(as.list(org.Hs.egREFSEQ2EG[x]), `[`, 1))

library(clusterProfiler)
library(ReactomePA)
xx <- compareCluster(clustEG, fun="enrichPathway",
                     organism="human", pvalueCutoff=0.05,
                     qvalueCutoff=0.05,
                     pAdjustMethod="none",
                     universe=unlist(clustEG))
# plot(xx, colorBy="qvalue")
dotplot(xx)
```

```{r reactome_summary, message=FALSE, warning=FALSE}
library(dplyr)
head(summary(xx) %>% dplyr::select(-geneID)) %>% kable()
```



